---
title: "BlackDataLabEdwardJonesProj"
author: "Bayowa"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

## D.C Black Population Count

library(tidyverse)
library(janitor)

# Read NHGIS File
blackpop_raw <- read_csv("Data_raw/nhgis0003_ts_nominal_tract.csv") %>%
  clean_names()

# Filter to DC
blackpop_dc <- blackpop_raw %>%
  filter(state == "District of Columbia")

# Select Black population variables 
pop_dc_black <- blackpop_dc %>%
  select(
    nhgiscode,
    tracta,
    name1970, name1980, name1990, name2000, name2010, name2020,
    b18ab1970,
    b18ab1980,
    b18ab1990,
    b18ab2000,
    b18ab2010,
    b18ab2020
  )

# Convert to long format
dc_black_long <- pop_dc_black %>%
  pivot_longer(
    cols = starts_with("b18ab"),
    names_to = "year",
    values_to = "black_population"
  ) %>%
  mutate(
    year = str_extract(year, "\\d{4}") %>% as.integer()
  )

# Attach correct tract name by decade
dc_black_long <- dc_black_long %>%
  mutate(
    tract_name = case_when(
      year == 1970 ~ name1970,
      year == 1980 ~ name1980,
      year == 1990 ~ name1990,
      year == 2000 ~ name2000,
      year == 2010 ~ name2010,
      year == 2020 ~ name2020
    )
  )

#Cleaned dataset
dc_black_population_clean <- dc_black_long %>%
  filter(!is.na(black_population)) %>%
  select(
    nhgiscode,
    tracta,
    tract_name,
    year,
    black_population
  ) %>%
  arrange(year, tracta)


#write_csv(
  #dc_black_population_clean,
  "Cleaned_data_EDJones/dc_black_population_tract_1970_2020.csv"
#)

View(dc_black_population_clean)

```

### Black Population %
```{r}
### 1960 Black (Non-White) Population
library(tidyverse)
library(janitor)

race_1960_raw <- read_csv("Data_raw/nhgis0002_ds92_1960_tract.csv") %>%
  clean_names()

# Filter to DC and select relevant columns
race_1960_dc <- race_1960_raw %>%
  filter(statea == "11") %>%            # FIPS code 11 = DC
  select(
    nhgiscode = gisjoin,
    tracta,
    areaname,
    total_population = b9t001,          # Total
    nonwhite_population = b7b003        # Non-White (Black proxy for 1960 DC)
  ) %>%
  mutate(
    year = 1960L,
    # Standardize TRACTA to match existing format
    tracta = str_pad(as.character(as.integer(tracta)), width = 4, pad = "0"),
    black_population = nonwhite_population,
    black_percent = black_population / total_population,
    tract_name = areaname
  ) %>%
  select(nhgiscode, tracta, tract_name, year, black_population,
         total_population, black_percent)


```



```{r}
library(tidyverse)
library(janitor)

race_raw <- read_csv("Data_raw/nhgis0003_ts_nominal_tract.csv") %>%
  clean_names()

race_dc <- race_raw %>%
  filter(state == "District of Columbia")

race_dc_selected <- race_dc %>%
  select(
    nhgiscode,
    tracta,
    name1970, name1980, name1990, name2000, name2010, name2020,

    # White
    b18aa1970, b18aa1980, b18aa1990, b18aa2000, b18aa2010, b18aa2020,

    # Black
    b18ab1970, b18ab1980, b18ab1990, b18ab2000, b18ab2010, b18ab2020,

    # AI/AN
    b18ac1970, b18ac1980, b18ac1990, b18ac2000, b18ac2010, b18ac2020,

    # Asian / Pacific Islander / Other
    b18ad1970, b18ad1980, b18ad1990, b18ad2000, b18ad2010, b18ad2020,

    # Two or more races (2000+)
    b18ae2000, b18ae2010, b18ae2020
  )

```


```{r}
race_long <- race_dc_selected %>%
  pivot_longer(
    cols = starts_with("b18"),
    names_to = "variable",
    values_to = "population"
  ) %>%
  mutate(
    year = str_extract(variable, "\\d{4}") %>% as.integer(),
    race = case_when(
      str_detect(variable, "b18aa") ~ "White",
      str_detect(variable, "b18ab") ~ "Black",
      str_detect(variable, "b18ac") ~ "American Indian or Alaska Native",
      str_detect(variable, "b18ad") ~ "Asian / Pacific Islander / Other",
      str_detect(variable, "b18ae") ~ "Two or More Races"
    )
  )
#Total population
dc_total_population_clean <- race_long %>%
  group_by(nhgiscode, tracta, year) %>%
  summarise(
    total_population = sum(population, na.rm = TRUE),
    .groups = "drop"
  )
#Black population
dc_black_population_clean <- race_long %>%
  filter(race == "Black") %>%
  mutate(
    tract_name = case_when(
      year == 1970 ~ name1970,
      year == 1980 ~ name1980,
      year == 1990 ~ name1990,
      year == 2000 ~ name2000,
      year == 2010 ~ name2010,
      year == 2020 ~ name2020
    )
  ) %>%
  select(
    nhgiscode,
    tracta,
    tract_name,
    year,
    black_population = population
  ) %>%
  filter(!is.na(black_population))
```

```{r}
dc_black_percent <- dc_black_population_clean %>%
  left_join(
    dc_total_population_clean,
    by = c("nhgiscode", "tracta", "year")
  ) %>%
  mutate(
    black_percent = black_population / total_population
  )
#write_csv(
  #dc_black_percent,
  #"Cleaned_data_EDJones/dc_black_population_percentage_1970_2020.csv"
#)
View(dc_black_percent)
```

```{r}
# Now bind 
dc_black_percent_full <- bind_rows(
  race_1960_dc,
  dc_black_percent        
) %>%
  arrange(nhgiscode, year)
View(dc_black_percent_full)
#write_csv(
  #dc_black_percent_full,
  #"Cleaned_data_EDJones/dc_black_population_percentage_1960_2020.csv"
#)
```



### Black Population Change
```{r}
dc_black_change <- dc_black_percent_full %>%
  arrange(nhgiscode, year) %>%
  group_by(nhgiscode) %>%
  mutate(
    black_pop_change = black_population - lag(black_population),
    black_pop_pct_change = (black_pop_change / lag(black_population)) * 100
  ) %>%
  ungroup()

#write_csv(
  #dc_black_change,
  #"Cleaned_data_EDJones/dc_black_population_change_1960_2020.csv"
#)
View(dc_black_change)
```

### Segregation
```{r}
dc_black_segregation <- dc_black_percent_full %>%
  mutate(
    segregation_category = case_when(
      black_percent < 0.25 ~ "0–25%",
      black_percent < 0.50 ~ "25–50%",
      black_percent < 0.75 ~ "50–75%",
      TRUE ~ "75–100%"
    )
  )
#write_csv(
  #dc_black_segregation,
  #"Cleaned_data_EDJones/dc_black_segregation_categories.csv"
#)
View(dc_black_segregation)
```

### Demographics (age,sex)
```{r}
library(tidyverse)
library(janitor)
library(jsonlite)

output_dir <- "Cleaned_data_EDJones"
if(!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

#Process 1960 Total Population (B8D) 
age_sex_1960_raw <- read_csv("Data_raw/nhgis0015_ds92_Age_sex_1960_tract.csv") %>%
  clean_names()

dc_age_sex_1960_clean <- age_sex_1960_raw %>%
  filter(statea == "11") %>%
  select(gisjoin, tracta, areaname, b8d001:b8d050) %>%
  pivot_longer(cols = b8d001:b8d050, names_to = "variable", values_to = "count") %>%
  mutate(
    # Refined sex logic: extract digits and check if odd (male) or even (female)
    var_num = as.numeric(str_extract(variable, "\\d+")),
    sex = if_else(var_num %% 2 == 1, "male", "female"),
    age_detailed = case_when(
      var_num %in% 1:2 ~ "Under 1 year",
      var_num %in% 3:4 ~ "1 year",
      var_num %in% 5:6 ~ "2 years",
      var_num %in% 7:8 ~ "3 years",
      var_num %in% 9:10 ~ "4 years",
      var_num %in% 11:12 ~ "5 years",
      var_num %in% 13:14 ~ "6 years",
      var_num %in% 15:16 ~ "7 years",
      var_num %in% 17:18 ~ "8 years",
      var_num %in% 19:20 ~ "9 years",
      var_num %in% 21:22 ~ "10 years",
      var_num %in% 23:24 ~ "11 years",
      var_num %in% 25:26 ~ "12 years",
      var_num %in% 27:28 ~ "13 years",
      var_num %in% 29:30 ~ "14 years",
      var_num %in% 31:32 ~ "15 years",
      var_num %in% 33:34 ~ "16 years",
      var_num %in% 35:36 ~ "17 years",
      var_num %in% 37:38 ~ "18 years",
      var_num %in% 39:40 ~ "19 years",
      var_num %in% 41:42 ~ "20 years",
      var_num %in% 43:44 ~ "21-24 years",
      var_num %in% 45:46 ~ "75-79 years",
      var_num %in% 47:48 ~ "80-84 years",
      var_num %in% 49:50 ~ "85 years and over",
      TRUE ~ NA_character_
    ),
    age_group = case_when(
      age_detailed %in% c("Under 1 year", "1 year", "2 years", "3 years", "4 years") ~ "Under 5 years",
      age_detailed %in% c("5 years", "6 years", "7 years", "8 years", "9 years") ~ "5 to 9 years",
      age_detailed %in% c("10 years", "11 years", "12 years", "13 years", "14 years") ~ "10 to 14 years",
      age_detailed %in% c("15 years", "16 years", "17 years") ~ "15 to 17 years",
      age_detailed %in% c("18 years", "19 years") ~ "18 and 19 years",
      age_detailed == "20 years" ~ "20 years",
      age_detailed == "21-24 years" ~ "22 to 24 years",
      age_detailed %in% c("75-79 years", "80-84 years") ~ "75 to 84 years",
      age_detailed == "85 years and over" ~ "85 years and over",
      TRUE ~ NA_character_
    ),
    year = 1960L,
    population = "total",
    source = "1960 Census - NT18/B8D"
  ) %>%
  filter(!is.na(age_group), count >= 0) %>%
  group_by(gisjoin, tracta, areaname, year, population, sex, age_group, source) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  # Standardize column names for binding later
  rename(geographic_id = gisjoin, tract_id = tracta, location_name = areaname) %>%
  mutate(tract_id = as.character(tract_id))

# Process 1970-2020 Total Population (B58)
b58_raw <- read_csv("Data_raw/nhgis0014_ts_nominal_Age_sex_1970_2020_tract.csv") %>%
  clean_names()

dc_b58 <- b58_raw %>%
  filter(statefp %in% c(11, "11") | state == "District of Columbia")

# Extract metadata for names
b58_tract_names <- dc_b58 %>%
  select(nhgiscode, tracta, starts_with("name")) %>%
  distinct()

dc_b58_long <- dc_b58 %>%
  select(nhgiscode, tracta, b58aa1970:b58bj2020) %>%
  pivot_longer(cols = b58aa1970:b58bj2020, names_to = "variable", values_to = "count") %>%
  mutate(
    sex = if_else(str_detect(variable, "b58a[a-r]"), "male", "female"),
    age_code = str_extract(variable, "(?<=b58)[a-z]{2}"),
    year = as.integer(str_extract(variable, "\\d{4}")),
    age_group = case_when(
      age_code %in% c("aa", "as") ~ "Under 5 years",
      age_code %in% c("ab", "at") ~ "5 to 9 years",
      age_code %in% c("ac", "au") ~ "10 to 14 years",
      age_code %in% c("ad", "av") ~ "15 to 17 years",
      age_code %in% c("ae", "aw") ~ "18 and 19 years",
      age_code %in% c("af", "ax") ~ "20 years",
      age_code %in% c("ag", "ay") ~ "21 years",
      age_code %in% c("ah", "az") ~ "22 to 24 years",
      age_code %in% c("ai", "ba") ~ "25 to 29 years",
      age_code %in% c("aj", "bb") ~ "30 to 34 years",
      age_code %in% c("ak", "bc") ~ "35 to 44 years",
      age_code %in% c("al", "bd") ~ "45 to 54 years",
      age_code %in% c("am", "be") ~ "55 to 59 years",
      age_code %in% c("an", "bf") ~ "60 and 61 years",
      age_code %in% c("ao", "bg") ~ "62 to 64 years",
      age_code %in% c("ap", "bh") ~ "65 to 74 years",
      age_code %in% c("aq", "bi") ~ "75 to 84 years",
      age_code %in% c("ar", "bj") ~ "85 years and over",
      TRUE ~ NA_character_
    )
  ) %>%
  left_join(b58_tract_names, by = c("nhgiscode", "tracta")) %>%
  mutate(
    location_name = case_when(
      year == 1970 ~ name1970,
      year == 1980 ~ name1980,
      year == 1990 ~ name1990,
      year == 2000 ~ name2000,
      year == 2010 ~ name2010,
      year == 2020 ~ name2020,
      TRUE ~ paste("Tract", tracta)
    ),
    population = "total",
    source = "NHGIS Time Series B58"
  ) %>%
  select(geographic_id = nhgiscode, tract_id = tracta, location_name, year, population, sex, age_group, count, source) %>%
  mutate(tract_id = as.character(tract_id))

# Combine and Finalize 
dc_age_sex_total_all <- bind_rows(dc_age_sex_1960_clean, dc_b58_long) %>%
  mutate(
    decade = paste0(floor(year / 10) * 10, "s"),
    data_quality = if_else(year == 1960, "limited - missing ages 25-64", "complete - B58 standardized")
  ) %>%
  arrange(year, tract_id, sex, age_group)

# Export & Summarize
write_csv(dc_age_sex_total_all, file.path(output_dir, "dc_age_sex_total_1960_2020.csv"))

total_sex_summary <- dc_age_sex_total_all %>%
  group_by(year, sex) %>%
  summarise(total_pop = sum(count, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = sex, values_from = total_pop)

print(total_sex_summary)
View(dc_age_sex_total_all)

```




### Income- Household income(NHGIS)
```{r}
library(tidyverse)
library(janitor)


# Create output directory
output_dir <- "Cleaned_data_EDJones"
if(!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

cat("Starting DC Income Data Processing (1960-2020)...\n")
cat("Output directory:", output_dir, "\n")


#PROCESS 1960 FAMILY INCOME DATA (Census)

cat("\nProcessing 1960 family income data...\n")

income_1960_raw <- read_csv("Data_raw/nhgis0008_ds92_1960HouseholdIncome_tract.csv") %>%
  clean_names()

# Filter for DC (STATEA = 11 for DC)
dc_income_1960 <- income_1960_raw %>%
  filter(statea == "11")

# Clean and prepare 1960 FAMILY income data
dc_income_1960_clean <- dc_income_1960 %>%
  select(
    gisjoin,
    year,
    tracta,
    areaname,
    # FAMILY income columns for 1959 (B8W table)
    b8w001:b8w013
  ) %>%
  # Convert to long format
  pivot_longer(
    cols = b8w001:b8w013,
    names_to = "variable_code",
    values_to = "families"
  ) %>%
  mutate(
    # Create meaningful income categories
    income_category = case_when(
      variable_code == "b8w001" ~ "<$1,000",
      variable_code == "b8w002" ~ "$1,000–1,999",
      variable_code == "b8w003" ~ "$2,000–2,999",
      variable_code == "b8w004" ~ "$3,000–3,999",
      variable_code == "b8w005" ~ "$4,000–4,999",
      variable_code == "b8w006" ~ "$5,000–5,999",
      variable_code == "b8w007" ~ "$6,000–6,999",
      variable_code == "b8w008" ~ "$7,000–7,999",
      variable_code == "b8w009" ~ "$8,000–8,999",
      variable_code == "b8w010" ~ "$9,000–9,999",
      variable_code == "b8w011" ~ "$10,000–14,999",
      variable_code == "b8w012" ~ "$15,000–24,999",
      variable_code == "b8w013" ~ "$25,000+"
    ),
    # Remove original variable code
    variable_code = NULL
  ) %>%
  # Add metadata
  mutate(
    unit = "families",  # Important: this is FAMILY income
    income_year = 1959,  # Income reported is for 1959
    data_type = "family_income",
    source = "1960 Census - Families by Income in 1959 (B8W table)",
    note = "1960 data reports FAMILY income, not household income. Filename is mislabeled."
  ) %>%
  rename(count = families)  # Standardize column name

cat(" 1960 data processed:", nrow(dc_income_1960_clean), "observations\n")


#PROCESS 1970-2000 HOUSEHOLD INCOME DATA (NHGIS Time Series)


cat("\nProcessing 1970-2000 household income data...\n")

income_raw <- read_csv(
  "Data_raw/nhgis0005_ts_houseincome_nominal_tract.csv",
  show_col_types = FALSE
) %>%
  clean_names()

income_dc <- income_raw %>%
  filter(state == "District of Columbia")

# Process 1970-2000 household income
dc_income_1970_2000 <- income_dc %>%
  select(
    nhgiscode,
    tracta,
    name1970, name1980, name1990, name2000,
    # Household income categories (4 brackets)
    bs7aa1970, bs7aa1980, bs7aa1990, bs7aa2000, # <$10k
    bs7ab1970, bs7ab1980, bs7ab1990, bs7ab2000, # $10k–14,999
    bs7ac1970, bs7ac1980, bs7ac1990, bs7ac2000, # $15k–24,999
    bs7ad1970, bs7ad1980, bs7ad1990, bs7ad2000  # $25k+
  ) %>%
  pivot_longer(
    cols = starts_with("bs7"),
    names_to = "variable",
    values_to = "households"
  ) %>%
  mutate(
    year = str_extract(variable, "\\d{4}") %>% as.integer(),
    income_category = case_when(
      str_detect(variable, "bs7aa") ~ "<$10,000",
      str_detect(variable, "bs7ab") ~ "$10,000–14,999",
      str_detect(variable, "bs7ac") ~ "$15,000–24,999",
      str_detect(variable, "bs7ad") ~ "$25,000+"
    ),
    tract_name = case_when(
      year == 1970 ~ name1970,
      year == 1980 ~ name1980,
      year == 1990 ~ name1990,
      year == 2000 ~ name2000
    )
  ) %>%
  filter(!is.na(households)) %>%
  select(
    nhgiscode,
    tracta,
    tract_name,
    year,
    income_category,
    households
  ) %>%
  # Add metadata
  mutate(
    unit = "households",
    income_year = year,
    data_type = "household_income",
    source = "NHGIS Time Series - Household Income (BS7 table)",
    note = "Nominal household income data. 4 income brackets."
  ) %>%
  rename(count = households)  # Standardize column name

cat(" 1970-2000 data processed:", nrow(dc_income_1970_2000), "observations\n")
cat("   Years:", paste(sort(unique(dc_income_1970_2000$year)), collapse = ", "), "\n")



# Read 2010 ACS data
acs_2010_raw <- read_csv("Data_raw/nhgis0009_ds191Household_MedianHouseincome_2008_20125_tract.csv") %>%
  clean_names()

# Filter for DC
dc_acs_2010 <- acs_2010_raw %>%
  filter(statea == "11")

# Process 2010 income distribution (QU0 = B19001 household income table)
dc_income_2010 <- dc_acs_2010 %>%
  select(
    gisjoin,
    tracta,
    name_e,
    # 16 income brackets from ACS table B19001
    qu0e002:qu0e017
  ) %>%
  pivot_longer(
    cols = qu0e002:qu0e017,
    names_to = "variable",
    values_to = "count"
  ) %>%
  mutate(
    # Map 16 ACS brackets to 4 categories for consistency
    income_category = case_when(
      variable == "qu0e002" ~ "<$10,000",
      variable == "qu0e003" ~ "$10,000–14,999",
      variable %in% c("qu0e004", "qu0e005") ~ "$15,000–24,999",
      TRUE ~ "$25,000+"  # Combine all higher brackets
    ),
    year = 2010,
    tract_name = name_e,
    unit = "households",
    income_year = 2012,  # ACS reports in 2012 inflation-adjusted dollars
    data_type = "household_income",
    source = "ACS 2008-2012 - Household Income (B19001 table)",
    note = "2012 inflation-adjusted dollars. 16 ACS brackets mapped to 4."
  ) %>%
  # Sum counts for each category (since we combined brackets)
  group_by(gisjoin, tracta, tract_name, year, income_category, 
           unit, income_year, data_type, source, note) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop")

cat(" 2010 data processed:", nrow(dc_income_2010), "observations\n")



# Read 2020 ACS data
acs_2020_raw <- read_csv("Data_raw/nhgis0009_ds262_Household_MedianHouseincome_2018_20225_tract.csv") %>%
  clean_names()

# Filter for DC
dc_acs_2020 <- acs_2020_raw %>%
  filter(statea == "11")

# Process 2020 income distribution (AQP5 = B19001 household income table)
dc_income_2020 <- dc_acs_2020 %>%
  select(
    gisjoin,
    tracta,
    name_e,
    # 16 income brackets from ACS table B19001
    aqp5e002:aqp5e017
  ) %>%
  pivot_longer(
    cols = aqp5e002:aqp5e017,
    names_to = "variable",
    values_to = "count"
  ) %>%
  mutate(
    # Map 16 ACS brackets to 4 categories for consistency
    income_category = case_when(
      variable == "aqp5e002" ~ "<$10,000",
      variable == "aqp5e003" ~ "$10,000–14,999",
      variable %in% c("aqp5e004", "aqp5e005") ~ "$15,000–24,999",
      TRUE ~ "$25,000+"  # Combine all higher brackets
    ),
    year = 2020,
    tract_name = name_e,
    unit = "households",
    income_year = 2022,  # ACS reports in 2022 inflation-adjusted dollars
    data_type = "household_income",
    source = "ACS 2018-2022 - Household Income (B19001 table)",
    note = "2022 inflation-adjusted dollars. 16 ACS brackets mapped to 4."
  ) %>%
  # Sum counts for each category (since we combined brackets)
  group_by(gisjoin, tracta, tract_name, year, income_category,
           unit, income_year, data_type, source, note) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop")

cat(" 2020 data processed:", nrow(dc_income_2020), "observations\n")



cat("\nStandardizing all datasets to common format...\n")

# Function to standardize column names and structure
standardize_income_dataset <- function(data) {
  cols <- names(data)
  
  # 1. Manually assign Geographic ID based on what exists
  if ("gisjoin" %in% cols) {
    data$geographic_id <- as.character(data$gisjoin)
  } else if ("nhgiscode" %in% cols) {
    data$geographic_id <- as.character(data$nhgiscode)
  } else {
    data$geographic_id <- NA_character_
  }
  
  #  Manually assign Location Name based on what exists
  if ("tract_name" %in% cols) {
    data$location_name <- as.character(data$tract_name)
  } else if ("areaname" %in% cols) {
    data$location_name <- as.character(data$areaname)
  } else if ("name_e" %in% cols) {
    data$location_name <- as.character(data$name_e)
  } else {
    data$location_name <- paste("Tract", data$tracta)
  }

  #  Now run the pipe for the logic that is consistent across all years
  data %>%
    mutate(
      tract_id = as.character(tracta),
      # Mapping to the 4-category broad schema for your longitudinal analysis
      income_category_broad = case_when(
        income_category %in% c("<$1,000", "$1,000–1,999", "$2,000–2,999",
                               "$3,000–3,999", "$4,000–4,999", "$5,000–5,999",
                               "$6,000–6,999", "$7,000–7,999", "$8,000–8,999",
                               "$9,000–9,999", "<$10,000") ~ "<$10,000",
        income_category == "$10,000–14,999" ~ "$10,000–14,999",
        income_category == "$15,000–24,999" ~ "$15,000–24,999",
        TRUE ~ "$25,000+"
      )
    ) %>%
    select(
      geographic_id,
      tract_id,
      location_name,
      year,
      income_category,
      income_category_broad,
      count,
      unit,
      income_year,
      data_type,
      source,
      note
    )
}

# Apply standardization
dc_income_1960_std <- standardize_income_dataset(dc_income_1960_clean)
dc_income_1970_2000_std <- standardize_income_dataset(dc_income_1970_2000)
dc_income_2010_std <- standardize_income_dataset(dc_income_2010)
dc_income_2020_std <- standardize_income_dataset(dc_income_2020)

# Combine All Decades
cat("\nCombining all decades (1960-2020)...\n")

dc_income_all_years <- bind_rows(
  dc_income_1960_std,
  dc_income_1970_2000_std,
  dc_income_2010_std,
  dc_income_2020_std
) %>%
  arrange(year, tract_id) %>%
  mutate(
    decade = paste0(floor(year / 10), "0s"),
    comparison_validity = case_when(
      year == 1960 & unit == "families" ~ "limited: family vs household",
      year >= 1970 & unit == "households" ~ "good: consistent unit",
      TRUE ~ "check unit consistency"
    )
  )

cat(" Combined dataset created:", nrow(dc_income_all_years), "total observations\n")

#  ADD INFLATION ADJUSTMENT FOR COMPARABILITY


cat("\nAdding inflation adjustment for cross-decade comparison...\n")

# CPI conversion factors (Consumer Price Index, 2020 base)
cpi_factors <- tibble(
  year = c(1960, 1970, 1980, 1990, 2000, 2010, 2020),
  cpi_2020 = c(9.21, 6.43, 2.91, 1.87, 1.48, 1.18, 1.00),
  inflation_base = c("1959 nominal", "1970 nominal", "1980 nominal",
                    "1990 nominal", "2000 nominal", "2012 adjusted",
                    "2022 adjusted"),
  cpi_source = "BLS CPI-U, annual average, 2020=1.00"
)

# Function to calculate bracket midpoints for analysis
calculate_bracket_midpoint <- function(income_category) {
  case_when(
    income_category == "<$1,000" ~ 500,
    income_category == "$1,000–1,999" ~ 1500,
    income_category == "$2,000–2,999" ~ 2500,
    income_category == "$3,000–3,999" ~ 3500,
    income_category == "$4,000–4,999" ~ 4500,
    income_category == "$5,000–5,999" ~ 5500,
    income_category == "$6,000–6,999" ~ 6500,
    income_category == "$7,000–7,999" ~ 7500,
    income_category == "$8,000–8,999" ~ 8500,
    income_category == "$9,000–9,999" ~ 9500,
    income_category == "<$10,000" ~ 5000,
    income_category == "$10,000–14,999" ~ 12500,
    income_category == "$15,000–24,999" ~ 20000,
    income_category == "$25,000+" ~ 37500,  # Approximate for distribution analysis
    TRUE ~ NA_real_
  )
}

# Apply inflation adjustment
dc_income_with_inflation <- dc_income_all_years %>%
  left_join(cpi_factors, by = "year") %>%
  mutate(
    # Calculate bracket midpoint in original dollars
    bracket_midpoint_original = calculate_bracket_midpoint(income_category),
    
    # Convert to 2020 dollars using CPI
    bracket_midpoint_2020 = bracket_midpoint_original * cpi_2020,
    
    # Academic documentation
    inflation_method = "CPI-U to 2020 dollars",
    adjustment_notes = paste(
      "Original data in", inflation_base, "dollars.",
      "Converted using CPI factor", round(cpi_2020, 2)
    )
  )

cat(" Inflation adjustment applied\n")

#
#  CALCULATE INCOME DISTRIBUTION METRICS


cat("\nCalculating income distribution metrics...\n")

dc_income_metrics <- dc_income_with_inflation %>%
  group_by(year, geographic_id, tract_id, location_name) %>%
  summarise(
    total_households = sum(count, na.rm = TRUE),
    
    # Income distribution percentages
    pct_under_10k = sum(count[income_category %in% c("<$1,000", "$1,000–1,999", "$2,000–2,999",
                                                   "$3,000–3,999", "$4,000–4,999", "$5,000–5,999",
                                                   "$6,000–6,999", "$7,000–7,999", "$8,000–8,999",
                                                   "$9,000–9,999", "<$10,000")]) / total_households * 100,
    pct_10k_to_15k = sum(count[income_category == "$10,000–14,999"]) / total_households * 100,
    pct_15k_to_25k = sum(count[income_category == "$15,000–24,999"]) / total_households * 100,
    pct_over_25k = sum(count[income_category == "$25,000+"]) / total_households * 100,
    
    # Income concentration (approximate)
    low_income_share = pct_under_10k + pct_10k_to_15k,
    middle_income_share = pct_15k_to_25k,
    high_income_share = pct_over_25k,
    
    .groups = "drop"
  ) %>%
  mutate(
    data_type = "income_distribution_metrics",
    source = "Calculated from income bracket data",
    note = "Percentages of households in each income category. For poverty rates, use official poverty tables."
  )

cat("Income metrics calculated\n")

# SAVE ALL OUTPUT FILES


cat("\nSaving output files...\n")

# Save complete income dataset
write_csv(
  dc_income_with_inflation,
  file.path("Cleaned_data_EDJones/dc_income_distributions_1960_2020.csv")
)
cat("dc_income_distributions_1960_2020.csv saved\n")

#  Save income metrics
write_csv(
  dc_income_metrics,
  file.path("Cleaned_data_EDJones/dc_income_metrics_1960_2020.csv")
) 
cat("Cleaned_data_EDJones/dc_income_metrics_1960_2020.csv saved\n")
View(dc_income_with_inflation)
View(dc_income_metrics)
#  Create and save metadata
metadata <- list(
  project_info = list(
    project_name = "DC Income Data for Edward P. Jones Literary Analysis",
    institution = "Black Data Lab Internship",
    completion_date = format(Sys.Date(), "%Y-%m-%d")
  ),
  
  data_sources = tribble(
    ~year, ~source, ~table_code, ~universe, ~dollar_type, ~key_notes,
    1960, "1960 Census", "B8W", "Families", "1959 nominal", "Family income (not households)",
    1970, "NHGIS Time Series", "BS7", "Households", "1970 nominal", "4 income brackets",
    1980, "NHGIS Time Series", "BS7", "Households", "1980 nominal", "4 income brackets",
    1990, "NHGIS Time Series", "BS7", "Households", "1990 nominal", "4 income brackets",
    2000, "NHGIS Time Series", "BS7", "Households", "2000 nominal", "4 income brackets",
    2010, "ACS 2008-2012", "B19001", "Households", "2012 adjusted", "16→4 brackets",
    2020, "ACS 2018-2022", "B19001", "Households", "2022 adjusted", "16→4 brackets"
  ),
  
  inflation_adjustment = list(
    method = "Consumer Price Index for All Urban Consumers (CPI-U)",
    base_year = 2020,
    source = "Bureau of Labor Statistics, Series CUUR0000SA0",
    factors = cpi_factors,
    purpose = "Enable valid dollar comparisons across decades"
  ),
  
  academic_notes = list(
    unit_caveat = "1960 data uses families while 1970-2020 uses households",
    bracket_mapping = "ACS 16-bracket data mapped to 4 brackets for consistency",
    comparison_recommendation = "Use percentages and inflation-adjusted values for cross-decade analysis",
    poverty_data = "Poverty rates should come from official poverty tables, not estimated from income distributions"
  )
)

# Save metadata as JSON
library(jsonlite)
write_json(
  metadata,
  file.path("Cleaned_data_EDJones/dc_income_metadata.json"),
  pretty = TRUE,
  auto_unbox = TRUE
)
cat(" dc_income_metadata.json saved\n")

# Create simple README
readme_content <- c(
  "DC INCOME DATA (1960-2020)",
  strrep("=", 60),
  "",
  "FILES:",
  "1. dc_income_distributions_1960_2020.csv - Complete income bracket data",
  "2. dc_income_metrics_1960_2020.csv - Calculated income distribution metrics",
  "3. dc_income_metadata.json - Complete documentation and methodology",
  "",
  "KEY FEATURES:",
  "- Covers 1960, 1970, 1980, 1990, 2000, 2010, 2020",
  "- Income data from Census, NHGIS, and ACS",
  "- Inflation-adjusted to 2020 dollars for comparison",
  "- Standardized 4-category income brackets",
  "",
  "ACADEMIC NOTES:",
  "- 1960: FAMILY income (different unit from later years)",
  "- 1970-2000: HOUSEHOLD income from NHGIS",
  "- 2010-2020: HOUSEHOLD income from ACS",
  "- Use percentages and inflation-adjusted values for comparisons",
  "",
  "CREATED:",
  paste("Date:", format(Sys.Date(), "%Y-%m-%d")),
  "Project: Black Data Lab Internship - Edward P. Jones Literary Analysis"
)

write_lines(readme_content, file.path("Cleaned_data_EDJones/DCINCOME_README.txt"))
cat("INCOME_README.txt saved\n")


# QUALITY CHECK AND SUMMARY


cat("\n")  
cat(strrep("=", 60))
cat("\n")
cat("PROCESSING COMPLETE: QUALITY SUMMARY\n")
cat("\n")
cat(strrep("=", 60))
cat("\n")

# Data coverage summary
coverage_summary <- dc_income_all_years %>%
  group_by(year, unit) %>%
  summarise(
    tracts = n_distinct(tract_id),
    total_observations = n(),
    total_households = sum(count, na.rm = TRUE),
    .groups = "drop"
  )

cat("DATA COVERAGE SUMMARY:\n")
print(coverage_summary)

# Check for missing data
missing_check <- dc_income_all_years %>%
  summarise(
    total_rows = n(),
    missing_count = sum(is.na(count)),
    zero_count = sum(count == 0, na.rm = TRUE),
    missing_pct = round(missing_count / total_rows * 100, 1)
  )

cat("\nDATA QUALITY CHECK:\n")
cat("Total rows:", missing_check$total_rows, "\n")
cat("Missing values:", missing_check$missing_count, "(", missing_check$missing_pct, "%)\n")
cat("Zero values:", missing_check$zero_count, "\n")

```

### Income- Poverty
```{r}
library(tidyverse)
library(janitor)

# Create output directory
output_dir <- "Cleaned_data_EDJones"
if(!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)


# PROCESS 2010 POVERTY DATA (ACS 2008-2012)

cat("Processing 2010 poverty data...\n")
poverty_2010_raw <- read_csv("Data_raw/nhgis0010_ds192_Poverty_lvl_2008_20125_tract.csv") %>%
  clean_names()

dc_poverty_2010_clean <- poverty_2010_raw %>%
  filter(statea == "11") %>%
  select(
    geographic_id = gisjoin,
    tract_id = tracta,
    location_name = name_e,
    poverty_universe = q9qe001,   # Total population for poverty determination
    below_poverty = q9qe002,      # Below poverty level
    above_poverty = q9qe031       # At or above poverty level
  ) %>%
  mutate(
    year = 2010,
    # Guard against division by zero in unpopulated tracts
    poverty_rate = ifelse(poverty_universe > 0, (below_poverty / poverty_universe) * 100, NA_real_),
    source = "ACS 2008-2012 - Poverty Status (B17001 table)",
    note = "5-year estimates. Population for whom poverty status determined.",
    data_type = "poverty_rate",
    geographic_id = as.character(geographic_id)
  )


#  PROCESS 2020 POVERTY DATA (ACS 2018-2022)

cat("Processing 2020 poverty data...\n")
poverty_2020_raw <- read_csv("Data_raw/nhgis0010_ds263_Poverty_lvl_2018_20225_tract.csv") %>%
  clean_names()

dc_poverty_2020_clean <- poverty_2020_raw %>%
  filter(statea == "11") %>%
  select(
    geographic_id = gisjoin,
    tract_id = tracta,
    location_name = name_e,
    poverty_universe = aq5ze001,
    below_poverty = aq5ze002,
    above_poverty = aq5ze031
  ) %>%
  mutate(
    year = 2020,
    poverty_rate = ifelse(poverty_universe > 0, (below_poverty / poverty_universe) * 100, NA_real_),
    source = "ACS 2018-2022 - Poverty Status (B17001 table)",
    note = "5-year estimates. Population for whom poverty status determined.",
    data_type = "poverty_rate",
    geographic_id = as.character(geographic_id)
  )


# PROCESS 1970-2000 POVERTY DATA (NHGIS Time Series)

cat("Processing 1970-2000 poverty data...\n")
poverty_raw <- read_csv("Data_raw/nhgis0006_ts_nominal_povertylvl_tract.csv", show_col_types = FALSE) %>%
  clean_names()

dc_poverty_1970_2000 <- poverty_raw %>%
  filter(state == "District of Columbia") %>%
  select(
    nhgiscode, tracta,
    name1970, name1980, name1990, name2000,
    ax6aa1970, ax6aa1980, ax6aa1990, ax6aa2000, # Universe
    cl6aa1970, cl6aa1980, cl6aa1990, cl6aa2000  # Below
  ) %>%
  pivot_longer(
    cols = starts_with(c("ax6", "cl6")),
    names_to = "variable",
    values_to = "population"
  ) %>%
  mutate(
    year = str_extract(variable, "\\d{4}") %>% as.integer(),
    measure = case_when(
      str_detect(variable, "ax6aa") ~ "poverty_universe",
      str_detect(variable, "cl6aa") ~ "below_poverty"
    ),
    location_name = case_when(
      year == 1970 ~ name1970,
      year == 1980 ~ name1980,
      year == 1990 ~ name1990,
      year == 2000 ~ name2000
    )
  ) %>%
  pivot_wider(id_cols = c(nhgiscode, tracta, location_name, year), 
              names_from = measure, values_from = population) %>%
  mutate(
    geographic_id = as.character(nhgiscode),
    tract_id = as.character(tracta),
    above_poverty = poverty_universe - below_poverty,
    poverty_rate = ifelse(poverty_universe > 0, (below_poverty / poverty_universe) * 100, NA_real_),
    source = "NHGIS Poverty Tables (1970-2000)",
    note = "Decennial census poverty data",
    data_type = "poverty_rate"
  ) %>%
  filter(!is.na(poverty_universe)) %>%
  select(geographic_id, tract_id, location_name, year, poverty_universe, 
         below_poverty, above_poverty, poverty_rate, source, note, data_type)


# COMBINE AND CATEGORIZE

cat("\nCombining all decades...\n")
dc_poverty_all_years <- bind_rows(
  dc_poverty_1970_2000,
  dc_poverty_2010_clean,
  dc_poverty_2020_clean
) %>%
  mutate(
    decade = paste0(floor(year / 10), "0s"),
    poverty_severity = case_when(
      is.na(poverty_rate) ~ "Missing data",
      poverty_rate < 10 ~ "Low poverty (<10%)",
      poverty_rate < 20 ~ "Moderate poverty (10-20%)",
      poverty_rate < 30 ~ "High poverty (20-30%)",
      poverty_rate >= 30 ~ "Very high poverty (30%+)"
    ),
    extreme_poverty_flag = ifelse(poverty_rate >= 40, 1, 0)
  )

# Calculate longitudinal changes within tracts
dc_poverty_final <- dc_poverty_all_years %>%
  group_by(geographic_id) %>%
  arrange(year) %>%
  mutate(
    poverty_rate_change = poverty_rate - lag(poverty_rate),
    pct_relative_change = ((poverty_rate - lag(poverty_rate)) / lag(poverty_rate)) * 100
  ) %>%
  ungroup()


#SAVE OUTPUTS

write_csv(dc_poverty_final, file.path(output_dir, "dc_poverty_1970_2020_cleaned.csv"))
cat("Saved: dc_poverty_1970_2020_cleaned.csv\n")

# Summary for console validation
summary_stats <- dc_poverty_final %>%
  group_by(year) %>%
  summarise(mean_rate = mean(poverty_rate, na.rm = TRUE), 
            tract_count = n(),
            .groups = "drop")
print(summary_stats)






```



### Education level
```{r}
library(tidyverse)
library(janitor)
library(jsonlite)

# Create output directory
output_dir <- "Cleaned_data_EDJones"
if(!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)


# PROCESS 1960 DATA

cat("Processing 1960 Education data...\n")
edu_1960_raw <- read_csv("Data_raw/nhgis0011_ds92_Education1960_tract.csv") %>% clean_names()
dc_edu_1960 <- edu_1960_raw %>% filter(statea == "11")

# Total Attainment (B6B)
dc_edu_1960_total <- dc_edu_1960 %>%
  select(gisjoin, tract_id = tracta, location_name = areaname, b6b001:b6b008) %>%
  pivot_longer(cols = b6b001:b6b008, names_to = "variable", values_to = "count") %>%
  mutate(
    attainment_broad = case_when(
      variable %in% c("b6b001", "b6b002", "b6b003", "b6b004", "b6b005") ~ "Less than high school",
      variable == "b6b006" ~ "High school graduate",
      variable == "b6b007" ~ "Some college/Associate's",
      variable == "b6b008" ~ "Bachelor's degree+"
    ),
    year = 1960, population = "total", source = "1960 Census (B6B)"
  ) %>%
  group_by(gisjoin, tract_id, location_name, year, population, attainment_broad, source) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop")

# Nonwhite Proxy (B8S)
dc_edu_1960_black <- dc_edu_1960 %>%
  select(gisjoin, tract_id = tracta, location_name = areaname, b8s001:b8s008) %>%
  pivot_longer(cols = b8s001:b8s008, names_to = "variable", values_to = "count") %>%
  mutate(
    attainment_broad = case_when(
      variable %in% c("b8s001", "b8s002", "b8s003", "b8s004", "b8s005") ~ "Less than high school",
      variable == "b8s006" ~ "High school graduate",
      variable == "b8s007" ~ "Some college/Associate's",
      variable == "b8s008" ~ "Bachelor's degree+"
    ),
    year = 1960, population = "black_proxy", source = "1960 Census (B8S)"
  ) %>%
  group_by(gisjoin, tract_id, location_name, year, population, attainment_broad, source) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop")


# PROCESS 1970 DATA

cat("Processing 1970 Education data...\n")
edu_1970_raw <- read_csv("Data_raw/nhgis0011_ds99_Education1970_tract.csv") %>% clean_names()
dc_edu_1970_total <- edu_1970_raw %>%
  filter(statea == "11") %>%
  select(gisjoin, tract_id = tracta, location_name = areaname, c2m001:c2m020) %>%
  pivot_longer(cols = c2m001:c2m020, names_to = "variable", values_to = "count") %>%
  mutate(
    attainment_broad = case_when(
      variable %in% c("c2m001", "c2m011", "c2m002", "c2m012", "c2m003", "c2m013", 
                       "c2m004", "c2m014", "c2m005", "c2m015", "c2m006", "c2m016") ~ "Less than high school",
      variable %in% c("c2m007", "c2m017") ~ "High school graduate",
      variable %in% c("c2m008", "c2m018") ~ "Some college/Associate's",
      variable %in% c("c2m009", "c2m019", "c2m010", "c2m020") ~ "Bachelor's degree+"
    ),
    year = 1970, population = "total", source = "1970 Census (C2M)"
  ) %>%
  group_by(gisjoin, tract_id, location_name, year, population, attainment_broad, source) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop")


# PROCESS 1980-2000 DATA (Race-aggregated)

# [Logic maintained from your draft, ensuring consistent names]
# Process 1980
edu_1980_raw <- read_csv("Data_raw/nhgis0011_ds107_Education1980_tract.csv") %>% clean_names()
dc_edu_1980_total <- edu_1980_raw %>%
  filter(statea == "11") %>%
  select(gisjoin, tract_id = tracta, location_name = areaname, dhn001:dhn025) %>%
  pivot_longer(cols = dhn001:dhn025, names_to = "variable", values_to = "count") %>%
  mutate(
    attainment_broad = case_when(
      str_detect(variable, "001|006|011|016|021|002|007|012|017|022") ~ "Less than high school",
      str_detect(variable, "003|008|013|018|023") ~ "High school graduate",
      str_detect(variable, "004|009|014|019|024") ~ "Some college/Associate's",
      str_detect(variable, "005|010|015|020|025") ~ "Bachelor's degree+"
    ),
    year = 1980, population = "total", source = "1980 Census (DHN)"
  ) %>%
  group_by(gisjoin, tract_id, location_name, year, population, attainment_broad, source) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop")


# PROCESS ACS DATA (2010 & 2020)

cat("Processing ACS 2010 & 2020 Education data...\n")
# 2010 Total
edu_2010_total <- read_csv("Data_raw/nhgis0013_ds191_Education2008_20125_tract.csv") %>%
  clean_names() %>% filter(statea == "11") %>%
  select(gisjoin, tract_id = tracta, location_name = name_e, qure003:qure035) %>%
  pivot_longer(cols = qure003:qure035, names_to = "variable", values_to = "count") %>%
  mutate(
    attainment_broad = case_when(
      variable %in% c("qure003", "qure004", "qure005", "qure006", "qure007", "qure008", "qure009", "qure010", 
                       "qure020", "qure021", "qure022", "qure023", "qure024", "qure025", "qure026", "qure027") ~ "Less than high school",
      variable %in% c("qure011", "qure028") ~ "High school graduate",
      variable %in% c("qure012", "qure013", "qure014", "qure029", "qure030", "qure031") ~ "Some college/Associate's",
      variable %in% c("qure015", "qure016", "qure017", "qure018", "qure032", "qure033", "qure034", "qure035") ~ "Bachelor's degree+"
    ),
    year = 2010, population = "total", source = "ACS 2008-2012"
  ) %>%
  group_by(gisjoin, tract_id, location_name, year, population, attainment_broad, source) %>%
  summarise(count = sum(count, na.rm = TRUE), .groups = "drop")


# COMBINE AND STANDARDIZE FIPS

dc_edu_total_all <- bind_rows(
  dc_edu_1960_total, dc_edu_1970_total, dc_edu_1980_total, 
  dc_edu_2010_total # Add your 1990/2000/2020 blocks here following same pattern
) %>%
  mutate(
    geographic_id = as.character(gisjoin),
    tract_id = as.character(tract_id),
    decade = paste0(floor(year / 10), "0s")
  )


# 6. SAVE OUTPUTS

write_csv(dc_edu_total_all, file.path(output_dir, "dc_education_total_1960_2020.csv"))
cat("✅ All education data processed and saved.\n")

```


# D.C Mortality as Health Indicator
```{r}
# 1960 mortality file
mort_1960_raw <- read_csv("Data_raw/nhgis0017_ds230_1960_mortality_county.csv") %>%
  clean_names()

# Filter DC
dc_1960 <- mort_1960_raw %>%
  filter(statea == "110") %>%
  select(year, total_deaths = agvv001)

#1960 Census population
dc_population_1960 <- 763956  #1960 DC population

dc_1960_clean <- dc_1960 %>%
  mutate(
    population = dc_population_1960,
    mortality_rate_per_100k = (total_deaths / population) * 100000,
    source = "NHGIS 1959-1972 Vital Statistics (NT003)",
    data_type = "crude_mortality_rate"
  )
View(dc_1960_clean)
```

```{r}
#mortality rate files
process_mortality <- function(file_path, year_value) {
  
  read_csv(file_path, show_col_types = FALSE) %>%
    clean_names() %>%
    filter(areaname == "District of Columbia") %>%
    mutate(
      mortality_rate_per_100k = agwh001 * 10,  # convert per 10k → per 100k
      year = year_value,
      source = "NHGIS 1970-2007 Vital Statistics (NT005)",
      data_type = "crude_mortality_rate"
    ) %>%
    select(year, mortality_rate_per_100k, source, data_type)
}

dc_1970 <- process_mortality("Data_raw/nhgis0017_ds231_1970_mortality_pertenthousand_county.csv", 1970)
dc_1980 <- process_mortality("Data_raw/nhgis0017_ds231_1980_mortality_county.csv", 1980)
dc_1990 <- process_mortality("Data_raw/nhgis0017_ds231_1990_mortality_county.csv", 1990)
dc_2000 <- process_mortality("Data_raw/nhgis0017_ds231_2000_mortality_county.csv", 2000)

nrow(dc_1970)
nrow(dc_1980)
nrow(dc_1990)
nrow(dc_2000)
```



```{r}
wonder_2010 <- read_csv("Data_raw/Underlying Cause of Death, 2010.csv", show_col_types = FALSE) %>%
  clean_names() %>%   
  # ADD THIS: filter out the footer notes from the CDC file
  filter(!is.na(county), county == "District of Columbia, DC") %>%
  mutate(
    year = 2010,
    total_deaths = deaths,
    mortality_rate_per_100k = crude_rate,
    source = "CDC WONDER - Underlying Cause of Death",
    data_type = "crude_mortality_rate"
  ) %>%
  select(
    year, total_deaths, population, mortality_rate_per_100k, source, data_type
  )

wonder_2020 <- read_csv(
  "Data_raw/Underlying Cause of Death, 2020-2.csv",
  show_col_types = FALSE
) %>%
  clean_names() %>%
  # FIX: changed 'County' to lowercase 'county' to match clean_names() output
  # Added !is.na to skip the 'Notes' rows at the bottom of the CDC file
  filter(!is.na(county), county == "District of Columbia, DC") %>%
  mutate(
    year = 2020,
    total_deaths = deaths,
    mortality_rate_per_100k = crude_rate,
    source = "CDC WONDER - Underlying Cause of Death",
    data_type = "crude_mortality_rate"
  ) %>%
  select(
    year,
    total_deaths,
    population,
    mortality_rate_per_100k,
    source,
    data_type
  )
View(wonder_2010)
View(wonder_2020)
```

```{r}
dc_1970_clean <- dc_1970 %>%
  mutate(
    total_deaths = NA,
    population = 756510
  ) %>%
  select(year, total_deaths, population, mortality_rate_per_100k, source, data_type)

dc_1980_clean <- dc_1980 %>%
  mutate(
    total_deaths = NA,
    population = 638334
  ) %>%
  select(year, total_deaths, population, mortality_rate_per_100k, source, data_type)

dc_1990_clean <- dc_1990 %>%
  mutate(
    total_deaths = NA,
    population = 606900
  ) %>%
  select(year, total_deaths, population, mortality_rate_per_100k, source, data_type)

dc_2000_clean <- dc_2000 %>%
  mutate(
    total_deaths = NA,
    population = 572056
  ) %>%
  select(year, total_deaths, population, mortality_rate_per_100k, source, data_type)
```


```{r}
dc_mortality_all <- bind_rows(
  dc_1960_clean,
  dc_1970_clean,
  dc_1980_clean,
  dc_1990_clean,
  dc_2000_clean,
  wonder_2010,
  wonder_2020
) %>%
  arrange(year)

View(dc_mortality_all)
write_csv(
  dc_mortality_all,
  "Cleaned_data_EDJones/dc_mortality_1960_2020.csv"
)
```

## Crime Rate D.C
```{r}
crime_raw <- read_csv(
  "Data_raw/crime rate D.C1960-2019.csv",
  skip = 1,      # adjust this number if needed
  col_names = c("year", "population", "index")
)

crime_clean <- crime_raw %>%
  filter(!is.na(year)) %>%
  mutate(
    year = as.integer(year),
    population = as.numeric(str_remove_all(population, ",")),
    index_rate_per_100k = as.numeric(str_remove_all(index, ","))
  ) %>%
  filter(year %in% c(1960, 1970, 1980, 1990, 2000, 2010)) %>%
  mutate(
    source = "FBI UCR Crime Index",
    data_type = "index_crime_rate"
  ) %>%
  select(year, population, index_rate_per_100k, source, data_type)
```

```{r}

crime_clean

```
#Crime Rate D.C Census Tract Variables
```{r}
# Load 2010
crime_2010 <- read_csv("Data_raw/Crime_Incidents_in_2010.csv") 
colnames(crime_2010)

```

```{r}
sum(is.na(crime_2010$CENSUS_TRACT))
length(unique(crime_2010$CENSUS_TRACT))
length(unique(crime_2010$WARD))
```


```{r}
sort(unique(crime_2010$OFFENSE))
```

```{r}
crime_2010_tract <- crime_2010 %>%
  filter(!is.na(CENSUS_TRACT)) %>%
  group_by(CENSUS_TRACT) %>%
  summarise(
    total_crime = n(),
    violent_crime = sum(OFFENSE %in% c(
      "HOMICIDE",
      "ROBBERY",
      "ASSAULT W/DANGEROUS WEAPON",
      "SEX ABUSE"
    )),
    property_crime = sum(OFFENSE %in% c(
      "BURGLARY",
      "MOTOR VEHICLE THEFT",
      "THEFT F/AUTO",
      "THEFT/OTHER",
      "ARSON"
    )),
    .groups = "drop"
  ) %>%
  mutate(
    year = 2010,
    geography = "tract",
    source = "DC Open Data Crime Incidents",
    data_type = "crime_count"
  )

glimpse(crime_2010_tract)
```

```{r}
sum(crime_2010_tract$violent_crime + crime_2010_tract$property_crime) ==
sum(crime_2010_tract$total_crime)
```


```{r}
sort(unique(crime_2010$WARD))
```



```{r}
# Load 2020
crime_2020 <- read_csv("Data_raw/Crime_Incidents_in_2020.csv") 
colnames(crime_2020)
```

```{r}
crime_2020_tract <- crime_2020 %>%
  filter(!is.na(CENSUS_TRACT)) %>%
  group_by(CENSUS_TRACT) %>%
  summarise(
    total_crime = n(),
    violent_crime = sum(OFFENSE %in% c(
      "HOMICIDE",
      "ROBBERY",
      "ASSAULT W/DANGEROUS WEAPON",
      "SEX ABUSE"
    )),
    property_crime = sum(OFFENSE %in% c(
      "BURGLARY",
      "MOTOR VEHICLE THEFT",
      "THEFT F/AUTO",
      "THEFT/OTHER",
      "ARSON"
    )),
    .groups = "drop"
  ) %>%
  mutate(
    year = 2020,
    geography = "tract",
    source = "DC Open Data Crime Incidents",
    data_type = "crime_count"
  )

glimpse(crime_2020_tract)
```

### Harmonize and create city-wide crimerate dataset for historical framing
```{r}
crime_2010_city <- crime_2010_tract %>%
  summarise(
    total_crime = sum(total_crime)
  ) %>%
  mutate(
    year = 2010,
    geography = "city",
    population = 604912,
    crime_rate_per_100k = (total_crime / population) * 100000,
    source = "DC Open Data Crime Incidents"
  )
```


```{r}
crime_2020_city <- crime_2020_tract %>%
  summarise(
    total_crime = sum(total_crime)
  ) %>%
  mutate(
    year = 2020,
    geography = "city",
    population = 712816,
    crime_rate_per_100k = (total_crime / population) * 100000,
    source = "DC Open Data Crime Incidents"
  )
```

```{r}
crime_clean <- crime_clean %>%
  mutate(
    geography = "city",
    total_crime = NA
  )
```


```{r}
crime_clean <- crime_clean %>%
  rename(
    crime_rate_per_100k = index_rate_per_100k
  )
```


```{r}
crime_clean <- crime_clean %>%
  select(
    year,
    population,
    crime_rate_per_100k,
    source,
    geography
  )
```

```{r}
crime_2010_city <- crime_2010_city %>%
  select(
    year,
    population,
    crime_rate_per_100k,
    source,
    geography
  )

crime_2020_city <- crime_2020_city %>%
  select(
    year,
    population,
    crime_rate_per_100k,
    source,
    geography
  )
```


```{r}
dc_crime_citywide <- bind_rows(
  crime_clean,
  crime_2010_city,
  crime_2020_city
) %>%
  arrange(year)

```

```{r}
dc_crime_citywide <- dc_crime_citywide %>%
  filter(!(year == 2010 & source == "DC Open Data Crime Incidents"))
```


```{r}
dc_crime_citywide <- dc_crime_citywide %>%
  arrange(year)
View(dc_crime_citywide)
```

```{r}
write_csv(
  dc_crime_citywide,
  "Cleaned_data_EDJones/dc_crime_citywide_1960_2020.csv"
)
```

### Harmonize and create tract-wide crime-rate dataset for granular visualization
```{r}
dc_crime_tract <- bind_rows(
  crime_2010_tract,
  crime_2020_tract
)
View(dc_crime_tract)
```


```{r}
write_csv(
  dc_crime_tract,
  "Cleaned_data_EDJones/dc_crime_tract_2010_2020.csv"
)
```


# Geographic ID Standardization

```{r}
library(tidyverse)
library(janitor)

output_dir <- "Cleaned_data_EDJones"

#pad tracta to 6 digits 
pad_tract <- function(x) {
  # Clean out decimals and scientific notation first
  clean <- format(x, scientific = FALSE, trim = TRUE)
  clean <- str_replace(clean, "\\.0$", "")
  # Pad to 6 digits
  str_pad(clean, width = 6, side = "left", pad = "0")
}

fix_fips_format <- function(x) {
  format(as.numeric(x), scientific = FALSE, trim = TRUE)
}

#Population percentage 
read_csv(file.path(output_dir, "dc_black_population_percentage_1960_2020.csv")) %>%
  rename(geographic_id = nhgiscode, tract_id = tracta, location_name = tract_name) %>%
  mutate(tract_join_id = pad_tract(tract_id)) %>%
  write_csv(file.path(output_dir, "dc_black_population_percentage_1960_2020_standard.csv"))


#Population change 
read_csv(file.path(output_dir, "dc_black_population_change_1960_2020.csv")) %>%
  rename(geographic_id = nhgiscode, tract_id = tracta, location_name = tract_name) %>%
  mutate(tract_join_id = pad_tract(tract_id)) %>%
  write_csv(file.path(output_dir, "dc_black_population_change_1960_2020_standard.csv"))


#Segregation 
read_csv(file.path(output_dir, "dc_black_segregation_categories.csv")) %>%
  rename(geographic_id = nhgiscode, tract_id = tracta, location_name = tract_name) %>%
  mutate(tract_join_id = pad_tract(tract_id)) %>%
  write_csv(file.path(output_dir, "dc_black_segregation_categories_standard.csv"))


#Age/sex 
read_csv(file.path(output_dir, "dc_age_sex_total_1960_2020.csv")) %>%
  mutate(tract_join_id = pad_tract(tract_id)) %>%
  write_csv(file.path(output_dir, "dc_age_sex_total_1960_2020_standard.csv"))


#Income distributions 
read_csv(file.path(output_dir, "dc_income_distributions_1960_2020.csv")) %>%
  mutate(tract_join_id = pad_tract(tract_id)) %>%
  write_csv(file.path(output_dir, "dc_income_distributions_1960_2020_standard.csv"))


# Income metrics 
read_csv(file.path(output_dir, "dc_income_metrics_1960_2020.csv")) %>%
  mutate(tract_join_id = pad_tract(tract_id)) %>%
  
 
  mutate(fips_full = fix_fips_format(tract_join_id)) %>% 
  
  write_csv(file.path(output_dir, "dc_income_metrics_1960_2020_standard.csv"))


#Education 
read_csv(file.path(output_dir, "dc_education_total_1960_2020.csv")) %>%
  mutate(tract_join_id = pad_tract(tract_id)) %>%
  write_csv(file.path(output_dir, "dc_education_total_1960_2020_standard.csv"))



#Crime tract file
read_csv(file.path(output_dir, "dc_crime_tract_2010_2020.csv")) %>%
  rename(tract_join_id = CENSUS_TRACT) %>%
  write_csv(file.path(output_dir, "dc_crime_tract_2010_2020_standard.csv"))



#Verification
cat("\n--- VERIFICATION: tract_join_id samples across files ---\n")
files_to_check <- c(
  "dc_black_population_percentage_1960_2020_standard.csv",
  "dc_age_sex_total_1960_2020_standard.csv",
  "dc_income_metrics_1960_2020_standard.csv",
  "dc_education_total_1960_2020_standard.csv",
  "dc_crime_tract_2010_2020_standard.csv"
)

for (f in files_to_check) {
  df <- read_csv(file.path(output_dir, f), show_col_types = FALSE)
  sample_ids <- head(unique(df$tract_join_id), 3)
  cat(sprintf("  %-55s tract_join_id samples: %s\n", f, paste(sample_ids, collapse = ", ")))
}
```


```{r}
library(tidyverse)

final_dir <- "Cleaned_data_EDJones/Final"
files_to_fix <- c(
  "dc_black_population_percentage_1960_2020_standard.csv",
  "dc_black_population_change_1960_2020_standard.csv",
  "dc_black_segregation_categories_standard.csv",
  "dc_education_total_1960_2020_standard.csv",
  "dc_income_metrics_1960_2020_standard.csv",
  "dc_crime_tract_2010_2020_standard.csv"
)

for (f in files_to_fix) {
  fpath <- file.path(final_dir, f)
  
  # Read EVERYTHING as character to prevent scientific notation
  df <- read_csv(fpath, col_types = cols(.default = "c")) %>%
    mutate(
      # Ensure fips_full is exactly 11 digits (state 11 + county 001 + 6-digit tract)
      # We use tract_join_id because it's already padded to 6 digits
      fips_full = paste0("11001", tract_join_id)
    )
  
  # Write it back
  write_csv(df, fpath)
  cat("Successfully locked FIPS as text in:", f, "\n")
}
```

```{r}
# Final Income Geometry Repair
dc_tracts_geo <- tracts(state = "DC", year = 2020, cb = TRUE) %>%
  mutate(fips_full = paste0(STATEFP, COUNTYFP, TRACTCE))

income_df <- read_csv(file.path(final_dir, "dc_income_metrics_1960_2020_standard.csv"), 
                      col_types = cols(fips_full = "c"))

income_geo <- income_df %>%
  left_join(dc_tracts_geo, by = "fips_full") %>%
  st_as_sf()

cat("Income Match Count:", sum(!is.na(st_dimension(income_geo))), "\n")
```


```{r}
library(tidyverse)

# Fix percentage file only
fpath <- file.path("Cleaned_data_EDJones", "dc_black_population_percentage_1960_2020_standard.csv")
read_csv(fpath, show_col_types = FALSE) %>%
  mutate(
    black_percent = if_else(is.na(black_percent) & total_population == 0, 0, black_percent)
  ) %>%
  write_csv(fpath)
cat("Fixed: dc_black_population_percentage_1960_2020_standard.csv\n")

# Fix change file only
fpath <- file.path("Cleaned_data_EDJones", "dc_black_population_change_1960_2020_standard.csv")
read_csv(fpath, show_col_types = FALSE) %>%
  mutate(
    black_percent = if_else(is.na(black_percent) & total_population == 0, 0, black_percent)
  ) %>%
  write_csv(fpath)
cat("Fixed: dc_black_population_change_1960_2020_standard.csv\n")

# Fix segregation file — this one has segregation_category too
fpath <- file.path("Cleaned_data_EDJones", "dc_black_segregation_categories_standard.csv")
read_csv(fpath, show_col_types = FALSE) %>%
  mutate(
    black_percent = if_else(is.na(black_percent) & total_population == 0, 0, black_percent),
    segregation_category = if_else(
      total_population == 0, "0-25%", segregation_category
    )
  ) %>%
  write_csv(fpath)
cat("Fixed: dc_black_segregation_categories_standard.csv\n")

```




```{r}
fpath <- file.path("Cleaned_data_EDJones", "dc_income_metrics_1960_2020_standard.csv")
read_csv(fpath, show_col_types = FALSE) %>%
  mutate(across(
    c(pct_under_10k, pct_10k_to_15k, pct_15k_to_25k, pct_over_25k,
      low_income_share, middle_income_share, high_income_share),
    ~ if_else(is.na(.) & total_households == 0, 0, .)
  )) %>%
  write_csv(fpath)
cat("Fixed: dc_income_metrics_1960_2020_standard.csv\n")
```

```{r}
library(tidyverse)
library(stringr)

input_dir <- "Cleaned_data_EDJones"
output_dir <- "Cleaned_data_EDJones/Final"

# Fix Income Distributions
read_csv(file.path(input_dir, "dc_income_distributions_1960_2020.csv"), show_col_types = FALSE) %>%
  mutate(
    # Create fips_full from geographic_id if fips_full doesn't exist yet
    fips_full = str_pad(as.character(geographic_id), width = 11, side = "left", pad = "0"),
    tract_join_id = str_pad(as.character(tract_id), width = 6, side = "left", pad = "0")
  ) %>%
  write_csv(file.path(output_dir, "dc_income_distributions_1960_2020_standard.csv"))

# Fix Income Metrics
read_csv(file.path(input_dir, "dc_income_metrics_1960_2020.csv"), show_col_types = FALSE) %>%
  mutate(
    fips_full = str_pad(as.character(geographic_id), width = 11, side = "left", pad = "0"),
    tract_join_id = str_pad(as.character(tract_id), width = 6, side = "left", pad = "0")
  ) %>%
  write_csv(file.path(output_dir, "dc_income_metrics_1960_2020_standard.csv"))

cat("Standardized files created with 11-digit FIPS codes.\n")
```


```{r}

```


```{r}
library(tidyverse)
library(janitor)
library(tigris)
library(sf)

# Check actual column names and sample values
crime_check <- read_csv(
  file.path(output_dir, "dc_crime_tract_2010_2020_standard.csv"),
  show_col_types = FALSE
)

cat("Crime tract columns:", names(crime_check), "\n")
cat("Sample tract_join_id values:\n")
print(head(crime_check$tract_join_id))

cat("\nPop file tract_join_id samples (2010):\n")
pop_check <- read_csv(
  file.path(output_dir, "dc_black_population_percentage_1960_2020_standard.csv"),
  show_col_types = FALSE
) %>% filter(year == 2010)

print(head(pop_check$tract_join_id))
cat("Match count:", sum(crime_check$tract_join_id %in% pop_check$tract_join_id), "\n")
```


```{r}
library(tidyverse)
library(tigris)
library(sf)

options(tigris_use_cache = TRUE)

dc_tracts_geo <- tracts(state = "DC", year = 2010, cb = FALSE) %>%
  select(fips_full = GEOID10, geometry) %>%
  mutate(fips_full = as.character(fips_full))

cat("Tract boundaries loaded:", nrow(dc_tracts_geo), "tracts\n")
print(head(dc_tracts_geo))
```

```{r}
output_dir <- "Cleaned_data_EDJones/Final"

# Files to add geometry to
tract_files <- c(
  "dc_black_segregation_categories_standard.csv",
  "dc_black_population_percentage_1960_2020_standard.csv",
  "dc_black_population_change_1960_2020_standard.csv",
  "dc_income_metrics_1960_2020_standard.csv",
  "dc_education_total_1960_2020_standard.csv",
  "dc_crime_tract_2010_2020_standard.csv"
)

for (fname in tract_files) {
  fpath <- file.path(output_dir, fname)
  
  # Read the data and FORCE 11-digit format (don't remove leading zero)
  df <- read_csv(fpath, show_col_types = FALSE) %>%
    mutate(fips_full = str_pad(as.character(fips_full), width = 11, side = "left", pad = "0"))
  
  # Join geometry on the matching 11-digit fips_full
  df_geo <- df %>%
    left_join(dc_tracts_geo, by = "fips_full") %>%
    st_as_sf()
  
  # Export as GeoJSON
  geojson_name <- str_replace(fname, "\\.csv$", ".geojson")
  geojson_path <- file.path(output_dir, geojson_name)
  
  # Only write if there's actually a geometry column to save
  st_write(df_geo, geojson_path, delete_dsn = TRUE, quiet = TRUE)
  
  # Verify match count
  matched <- sum(!is.na(st_dimension(df_geo)))
  cat(sprintf(" %s — %d rows matched to geometry\n", geojson_name, matched))
}
  
```


```{r}
# Check your CSV format (should be 11 digits)
head(df$fips_full)

# Check the Shapefile format (this is the mystery guest)
head(dc_tracts_geo$fips_full)
```


```{r}
library(sf)
library(tidyverse)
library(tigris)
options(tigris_use_cache = TRUE)

output_dir <- "Cleaned_data_EDJones/Final"
shapefile_dir <- file.path(output_dir, "shapefiles")
dir.create(shapefile_dir, showWarnings = FALSE)

# Get the geometry (Tigris returns 11-digit FIPS by default)
dc_tracts_geo <- tracts(state = "DC", year = 2020, cb = TRUE) %>%
  mutate(fips_full = paste0(STATEFP, COUNTYFP, TRACTCE)) %>%
  select(fips_full) # geometry is included automatically

for (fname in geo_files) {
  fpath <- file.path(output_dir, fname)
  layer_name <- str_replace(fname, "\\.csv$", "")
  
  # Read and standardize FIPS to 11 digits to match Tigris
  df <- read_csv(fpath, show_col_types = FALSE) %>%
    mutate(fips_full = str_pad(as.character(fips_full), width = 11, side = "left", pad = "0"))
  
  # Join
  df_geo <- df %>%
    left_join(dc_tracts_geo, by = "fips_full") %>%
    st_as_sf()
  
  # Final verification check before writing
  matched_count <- sum(!is.na(st_dimension(df_geo)))
  
  out_path <- file.path(shapefile_dir, paste0(layer_name, ".shp"))
  st_write(df_geo, out_path, delete_dsn = TRUE, quiet = TRUE)
  
  cat(sprintf(" %s — %d rows matched to geometry\n", layer_name, matched_count))
}

cat("\nDone! All shapefiles are ready for ArcGIS in Final/shapefiles/\n")
```



```{r}
# Read the income data
df_income <- read_csv(file.path(output_dir, "dc_income_metrics_1960_2020_standard.csv"), show_col_types = FALSE)

# Advanced ID Repair
df_fixed <- df_income %>%
  mutate(
    # Clean the tract_id (remove decimals if they exist)
    clean_tract = str_replace(as.character(tract_id), "\\.", ""),
    # Modern tracts are 6 digits. 
    # If it's "1", it needs to be "000100".
    # If it's "2201", it needs to be "002201".
    tract_6_digit = str_pad(clean_tract, width = 6, side = "right", pad = "0"),
    tract_6_digit = str_pad(tract_6_digit, width = 6, side = "left", pad = "0"),
    
    # Reconstruct the 11-digit FIPS: 11 (DC) + 001 (DC County) + 6-digit tract
    fips_full = paste0("11001", tract_6_digit)
  )

# Join with your 2020 boundaries
df_geo <- df_fixed %>%
  left_join(dc_tracts_geo, by = "fips_full") %>%
  st_as_sf()

# Final verification
matched_count <- sum(!is.na(st_dimension(df_geo)))
cat(sprintf("🎯 Progress: %d rows matched! (Goal is ~1,100+)\n", matched_count))

# Overwrite the shapefile
st_write(df_geo, file.path(shapefile_dir, "dc_income_metrics_1960_2020_standard.shp"), delete_dsn = TRUE, quiet = TRUE)
```



```{r}
library(tidyverse)

geo_files <- c(
  "dc_black_population_change_1960_2020_standard.csv",
  "dc_black_population_percentage_1960_2020_standard.csv",
  "dc_black_segregation_categories_standard.csv",
  "dc_crime_tract_2010_2020_standard.csv",
  "dc_education_total_1960_2020_standard.csv",
  "dc_income_metrics_1960_2020_standard.csv"
)

for (fname in geo_files) {
  df <- read_csv(file.path("Cleaned_data_EDJones/Final", fname), 
                 show_col_types = FALSE, n_max = 1)
  cat(sprintf("\n%s:\n", fname))
  print(names(df))
}
```



